Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-02-24 draft.tex
.

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
A key step in characterizing a control loop in the context of biological systems should be to identify the previously described compartments. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and "straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. Firing rate is - as stated earlier - an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a "sensible" quantity. In a nerve cell this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section} and \ref{NO_Experiments_Section}. Given a representation of activity, its deviation from a reference value will cause other physiological compartments to affect the neuron's ability to generate action potentials. Referring to the concepts introduced in the previous section one could then ask what type of control best resembles the actual neuronal control. The fact that neurons, whose firing rate was artificially increased or suppressed over a long time, can reliably tune their activity back to the initial state (see e.g. \cite{Burrone_2002}) rules out a purely differential type of control. The return to baseline under different disturbances seems to point towards an integral-like control. It is clear however that a the ability of a biological system to encode the accumulation of an error signal is limited by the finite amount of available resources. The conductance of an ion channel for example has natural upper and lower bounds that it can not surpass. We will pick up this issue in Section \ref{Possible_Modifications_Section}.

@@

Another problem that one might consider is the fact that temporal fluctuations of activity are believed to be a major part of information transmission and processing within the brain \cite[p. 123-150]{Theor_Neur_Dayan}. Homeostasis as described above could potentially override information by eliminating these fluctuations of activity. However, it is argued that homeostatic mechanisms act on time scales of hours to days and thus do not interfere with changes in activity taking place on the order of seconds or below \cite{Turrigiano_2011}. Returning to the previous question, though the slowness of homeostatic adaptation might allow for temporal fluctuations of activity, it does not overcome the conceptual dichotomy between single cell homeostasis and persistent variability in spontaneous activity. Marder and Goaillard speculated that instead of a mechanism controlling activity on a single-cell level, one might consider the possibility of a "global sensor of network performance to provide error signals for tuning of the network" \cite{Marder_2006}. This concept was used by Sweeney et al. as the basic idea for the introduction of a diffusive, intercellular homeostatic mechanism, which will be introduced in the next section.

@@

\textbf{Synaptic Pruning:} At the same rate of 1/sec, EE synapses below a threshold of 0.000001 mV were removed, thus being added again to the set of "potential" connections from which the growth process draws new connections.

\textbf{Spike Timing Dependent Plasticity:} An additive STDP rule was used as described e.g. in \cite{Zhang_STDP}. The change of weight between two neurons due to a pre- and postsynaptic spike (i $\rightarrow$ j) is defined as:
\begin{align}
\Delta w_{ji} &= \sum_k \sum_l W(t_j^l - t_i^k) \label{STDP_rule} \\
W(\Delta t) &= A_{+} exp(-\Delta t / \tau_{+}), & \Delta t > 0 \label{STDP_pos} \\
W(\Delta t) &= A_{-} exp(\Delta t / \tau_{-}), & \Delta t < 0 \label{STDP_neg}
\end{align}
Indexes k and l refer to the k-th and l-th pre- and postsynaptic spike respectively. Parameters were chosen to approximate data from \cite{Bi_Poo_STDP} and \cite{Froemke_STDP}, namely $\tau_{+} = 15 ms$, $A_{+} = 15 mV$, $\tau_{-} = 30 ms$ and $A_{-} = -7.5 mV$. However, we used the "nearest neighbor" approximation for the sake of reduction of computational effort, only calculating the effect of the most recent pre-post pair of spikes for potentiation and post-pre pair for depression, yielding roughly the same value as the full summation due to the fast decay times $\tau_{+}$ and $\tau_{-}$ of the STDP-window.

@@

The new diffusive homeostatic model by Sweeney et al. combined the empirical findings discussed in Section \ref{NO_Experiments_Section} into a set of differential equations:   
\begin{align}
\dot{{Ca^{2+}}_i}(t) &= -\frac{{Ca^{2+}}_i}{\tau_{Ca^{2+}}} + Ca^{2+}_{spike} \sum_{j} \delta(t-t_{spike,i,j}) \label{Ca_dyn}\\
\dot{nNOS_i}(t) &= \frac{1}{\tau_{nNOS}} \left(\frac{{{Ca^{2+}}_i}^3}{{{Ca^{2+}}_i}^3+1} - nNOS_i \right) \label{nNOS_dyn}\\
\dot{NO}(\mathbf{r},t)&=-\lambda NO + D \nabla^2 NO + \sum_{i} \delta^2(\mathbf{r}-\mathbf{r}_{neur,i})\cdot nNOS^i \label{NO_dyn}\\
\dot{V_{t,i}}(t) &= \frac{NO(\mathbf{r}_{neur,i},t)-NO_0}{NO_0\cdot\tau_{V_t}} \label{Theta_dyn}
\end{align}

A depolarization within a nerve cell upon a spike-event $t_{spike}$ causes a fixed inflow of ionic current $Ca^{2+}_{spike}$, which is modeled as an instantaneous increase of the $Ca^{2+}$ concentration. The concentration decays exponentially by a time constant $\tau_{Ca^{2+}}$, see \eqref{Ca_dyn}. Though $Ca^{2+}$ currents can be described in a much more detailed fashion, it can be considered as a reasonable approximation \cite[p.~198-203]{Theor_Neur_Dayan}. The influence of $Ca^{2+}$ onto nNOS was modeled by Sweeney et al. through \eqref{nNOS_dyn} using the Hill equation \cite{Hill_Equ} to model a cooperative binding mechanism. The $nNOS$ production is then fed into the "pool" of nitric oxide via point sources located at the neurons' positions. An additional decay term was added apart from the inflow and the diffusive term to provide a stable finite $NO$ concentration under constant neuronal activity.

@@

The appropriate choice of $NO_0$ is obviously crucial for the goal of achieving and maintaining a certain level of activity. However, one cannot directly set a parameter of the model to the desired population activity as it was the case for canonical intrinsic homeostasis. One rather needs to determine the average concentration \textit{associated} with the desired activity and set it as a target concentration. Though it is possible to derive this relation in an analytic fashion, we let the system run with the previous homeostatic mechanism, still solving equation \eqref{Ca_dyn}-\eqref{NO_dyn} until a steady mean over the concentrations at the neurons' positions was reached. This mean was then set to be the target concentration and we switched to diffusive homeostasis. Table \ref{Params_IP} summarizes the choice of parameters that were introduced in this section. Diffusion parameters roughly match those measured in experiments \cite{Philippides_2000}.
\begin{table}
\begin{tabular}{|l|l|}
\hline
\textbf{parameter} & \textbf{value} \\
\hline
$\mathrm{r_{IP}}$ & $\mathrm{3\;Hz}$ \\
\hline
$\mathrm{\eta_{IP}}$ & $\mathrm{0.1\;mV}$ \\
\hline
$\mathrm{Ca^{2+}_{spike}}$ & 1 \\ \hline
$\mathrm{\tau_{Ca^{2+}}}$ &  $\mathrm{10\;ms}$ \\
\hline
$\mathrm{\tau_{nNOS}}$ & $\mathrm{100\;ms}$ \\
\hline
$\mathrm{D}$ & default: 10 $\mathrm{\mu m^2 ms^{-1}}$ \\
\hline 
$\mathrm{\lambda}$ & $\mathrm{0.1\;s^{-1}}$ \\
\hline
$\mathrm{\tau_{V_t}}$ & see section \ref{activ_analys} \\
\hline
\end{tabular}
\caption{Parameters of homeostatic intrinsic plasticity.}
\label{Params_IP}
\end{table}

@@

3.) Dirichlet boundary conditions with $NO = NO_{bound}$ at the boundaries.\\
Neumann boundary conditions were used for most of the simulations if not explicitly marked differently. This decision relates to the previously described mechanism of synaptic growth: Neurons placed close to the edge of the sheet have a lower connection probability due to the absence of neighboring neurons in the direction perpendicular to the close-by border. It therefore models the synaptic growth within a square "cutout" of neural tissue. The Neumann boundary condition fits into this picture since it allows a zero-flux condition at the borders. This is a reasonable assumption, because NO molecules cannot diffuse out of the tissue (unless they were placed in a fluid surrounding).

Equation \eqref{NO_dyn} describes the influx of NO as a sum of scaled and spatially shifted Dirac functions. Apart from the question whether this source term results in a well defined, finite solution at the neurons' positions (see section \ref{Matrix_Diff_Model_Section}), it can only be modeled to a certain degree of accuracy depending on the resolution of the numeric grid. In practice, we approximated the point sources of NO as an insertions at individual grid cells at a rate of $nNOS_i(t) / h^2$, where the normalizing divisor $h^2$ ensured the desired total influx per neuron. This numeric implementation required two additional constraints: First, all random neuron positions were confined to integer multiples of $h$ in x- and y-direction. Second, to avoid redundancy and for physiological reasons, each grid cell could only hold one neuron at maximum. Both conditions combined led to an iterative generation of positions where for each neuron a random number generator produced pairs of integers $(n_x,n_y)$, each within $[0,N]$, until an unoccupied pair of integers was found and occupied, moving on to the next neuron. Figure \ref{diff_test_plot} shows an example of the resulting NO density throughout a simulation. 
\begin{figure}[h!]
%% density_plot.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/density_plots/density_plot.png}
\end{center}
\caption{Example of NO-diffusion with 400 point sources of excitatory neurons.}
\label{diff_test_plot}
\end{figure}

\newpage
\section{Results} \label{results}
We first present the results of a comparison between features of the network activity in both variants in section \ref{activ_analys}, since our main goal of the implementation of diffusive homeostasis was to allow the network to develop a broader distribution of firing rates across excitatory neurons compared to the original version of homeostasis. This is followed by an analytic discussion of an instability we observed within the diffusive homeostatic feedback loop, see section \ref{theor_osc}. Furthermore, we compare topological features of the network under the influence of diffusive and non-diffusive homeostasis in section \ref{Network_Topology_Section}.

@@

where $\tau_e$ and $\tau_i$ are time constants representing the rapidity of firing rate adaptation, $\phi_e$ and $\phi_i$ the respective transition functions between synaptic input and firing rate and $W_{xy}$ the mean synaptic weight from population x to population y. A persistent oscillation then corresponds to a stable limit cycle in the 2-dimensional phase plane. Such a limit cycle exists inside a region $R$ if $R$ contains no fixed points and if any trajectory whose starting points lies within $R$ is confined to $R$, according to the Poincaré-Bendixson theorem. The second condition is equivalent to the condition that for all points $\mathbf{r}_{edge}$ on the edge of $R$, the vector field $\dot{\mathbf{r}}_{edge}$ is facing "into" $R$, see \cite[p.~248]{Dyn_Sys_Hirsch}. This case occurs if the system has a locally unstable fixed point, but a non-linearity that prevents an infinite deviation from the fixed point.

@@

{Ca^{2+}}^3(t) &= \left[ Ca^{2+}_{spike} \sum_{i} \theta(t-t^i_{spike,i}) exp(-(t-t^i_{spike,j})/\tau_{Ca^{2+}}) \right]^3 \\
&\approx {Ca^{2+}_{spike}}^3 \sum_{i} \theta(t-t^i_{spike,i}) exp(-3(t-t^i_{spike,j})/\tau_{Ca^{2+}})

@@

Returning to the question of how to describe NO synthesis in a simplified form, we note that according to equation \eqref{nNOS_single_sol_approx2}, a single spike causes the release of NO by an amount of $\frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$, which makes the mean rate of NO production over time simply $<nNOS> = <r> \cdot \frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$. However, it is not sufficient to simply propose $nNOS_i(t) = r_i(t) \cdot \frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$. A single cell fires in the range of $\mathrm{3\,Hz}$, which is - at least on the time scale of the observed oscillations - not enough to assign an "instantaneous" firing rate.
\newpage
We introduced an idealized picture of diffusive homeostasis to deal with this, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses $nNOS_{total}(t) \equiv \sum_{i} nNOS^i(t)$ and the decay term $-\lambda NO$. , $nNOS_{total}$ must be divided by the area $L^2$ of the tissue to account for the spread of NO. Furthermore, we introduce a random fluctuation term $\sigma_{NO} \xi(t)$ that accounts for local, momentary deviations from the mean. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{nNOS_{total}(t)}{L^2} + \sigma_{NO} \xi(t)\;.
\label{NO_dyn_approx}
\end{equation}
The amount of noise in the system is a hard-to-predict quantity since it depends on the spatial structure on the diffusive lattice, giving neurons that are further separated less impact onto each other. We tried to estimate the noise level based on the same instant-spread-simplification made in Equation \eqref{NO_dyn_approx} to describe the overall NO-dynamics. This led to an underestimation of the total energy within the power spectrum in the final result. However, noise level only acts as a global scaling factor of the shape of the power spectrum as long as we assume the noise in the system to be approximately white and was thus left open as a model fitting parameter.

@@

Figure \ref{Pow_Spec_Theta_vs_Analytic} shows the result of \eqref{Pow_Spec_Theta_Final}. Choosing noise amplitude to be the only free parameter of the model gave a rather unsatisfying result. Our approximation underestimated the amount of damping within the system (which controls the "peakiness" of the spectrum), as well as predicting a slightly smaller preferred amplitude. The preferred amplitude is controlled by $\frac{\gamma \alpha N_{exc.}}{L^2}$ and $\frac{1}{NO_0\tau_{V_t}}$ while the broadness of the spectrum is mostly controlled by the damping term $-\lambda$. Since the latter term was taken unaltered from the original system (apart from a shift in phase space), we added $-\lambda$ and $\alpha$ as free fitting parameters and managed to get a very good fit to the simulation data, see Figure \ref{Pow_Spec_Theta_vs_Analytic}.   
\begin{figure}
%% plot_pow_spec_analytic.py
\begin{center}
\includegraphics[width=\textwidth]{../../plots/power_spec/power_spectrum_analytic.png}
\end{center}
\caption{Power Spectrum of $\mathrm{\langle V_t \rangle}$ (blue) and analytic predictions based on Equation \eqref{Pow_Spec_Theta_Final}. Red curve was fitted by tweaking only the overall amplitude $\mathrm{\sigma_{NO}}$ in \eqref{Pow_Spec_Theta_Final} ($\mathrm{\alpha = -0.789 \,Hz/mV}$, as in Figure \ref{thresh_r_linfit}). Green curve is a fit achieved by adjusting $\mathrm{\sigma_{NO}}$ and $\mathrm{\lambda}$ in \eqref{Pow_Spec_Theta_Final} ($\mathrm{\lambda = 0.845 \,s^{-1}}$). Black curve was fitted by means of $\mathrm{\sigma_{NO}}$, $\mathrm{\lambda}$ and $\mathrm{\alpha}$ ($\mathrm{\lambda = 0.592 \,s^{-1}}$, $\mathrm{\alpha=-0.989 \,Hz/mV}$).}
\label{Pow_Spec_Theta_vs_Analytic}
\end{figure}
The fact that the value for $\mathrm{\alpha}$ that was determined by means of the linear fit in Figure \ref{thresh_r_linfit} leads to an underestimation of the preferred frequency can be partially explained by the linear regression, which overestimates the importance of a number of high-threshold-outliers. This leads to a shallower slope than what would be appropriate for the "working point" of oscillatory behavior. Another possible explanation could be that "instant spread" is only an approximation. Rather, one could picture the nitrous oxide just to fill a fraction of the available space on a short time scale by replacing $L^2$ by a smaller area. This leads to a stronger local increase of concentration, which eventually leads to faster oscillations. In either case, both parameters have the same effect onto the shape of the power spectrum since these parameters appear as a fraction $\alpha/L^2$ in \eqref{Pow_Spec_Theta_Final}.

@@

From a more general perspective one could argue that a mechanism binding firing rates to spatial constraints suppresses the ability of the system to adapt to a changing environment since it essentially "freezes" the state of activity. Experiments have shown that neurons' firing rates are indeed correlated under a changing environment/task: Neurons with a certain baseline-firing rate are very likely to have a similar firing rate while performing a task \cite{Buzsaki_Fir_Rates_2014}. While homeostatic fixation might not be problematic in the case of transient changes of stimuli, a spatial fixation of firing rates would presumably contradict the fact that long-term learning processes and plasticity lead to persistent changes in activity patterns \cite{Lever_Long_Term_Plast_2002}.

@@

Taking a closer look at the distributions depicted in Figure \ref{Weight_Dist} we noted that the increased width of the distribution is largely due to to an increment on the left half of the curve. Since new weights are initialized at $\mathrm{10^{-4}\, mV}$ this might indicate that some weights are drifting much more slowly from their value of initialization compared to other weights. These "slow drifters" are presumably connecting neurons both spiking at a below-average firing rate.
This provides an explanation why this effect is more present in the case of diffusive homeostasis since it causes a portion of neurons to have rather low firing rates. On the other hand, for non-diffusive homeostasis, one can assume that the amount of STDP-fluctuations is approximately equal for all recurrent excitatory weights because of the fixation of firing rates.

>>>>

\subsubsection{Error Signals and Controllers of Neuronal Homeostasis}
A key step in characterizing a control loop in the context of biological systems should be to identify the previously described compartments. While the functional separation illustrated in Figure \ref{Closed_Loop_Control_Illustration} might seem intuitive and ``straight forward", it is not obvious which physiological element should be attributed to a certain abstract element.
As a start, we state that the process and its output controlled by homeostasis is the generation of action potentials and the resulting firing rate. Firing rate is - as stated earlier - an abstract quantity, requiring a feedback transducer to translate the neuron's spiking activity into a ``sensible" quantity. In a nerve cell this could be represented by e.g. the relation between the firing of action potentials and the temporary rise in $\mathrm{Ca^{2+}}$ concentration. An example for a homeostatic model making use of intracellular calcium concentration as a measure of spiking activity is the LeMasson-Abbott model \cite{LeMasson_1993}. More complex chains of activity encoding are of course possible and we will introduce a recent model making additional use of the concentration of nitric oxide in Section \ref{DIff_Hom_Intro_Section} and \ref{NO_Experiments_Section}. Given a representation of activity, its deviation from a reference value will cause other physiological compartments to affect the neuron's ability to generate action potentials. Referring to the concepts introduced in the previous section one could then ask what type of control best resembles the actual neuronal control. The fact that neurons, whose firing rate was artificially increased or suppressed over a long time, can reliably tune their activity back to the initial state (see e.g. \cite{Burrone_2002}) rules out a purely differential type of control. The return to baseline under different disturbances seems to point towards an integral-like control. It is clear however that a the ability of a biological system to encode the accumulation of an error signal is limited by the finite amount of available resources. The conductance of an ion channel for example has natural upper and lower bounds that it can not surpass. We will pick up this issue in Section \ref{Possible_Modifications_Section}.

@@

Another problem that one might consider is the fact that temporal fluctuations of activity are believed to be a major part of information transmission and processing within the brain \cite[p. 123-150]{Theor_Neur_Dayan}. Homeostasis as described above could potentially override information by eliminating these fluctuations of activity. However, it is argued that homeostatic mechanisms act on time scales of hours to days and thus do not interfere with changes in activity taking place on the order of seconds or below \cite{Turrigiano_2011}. Returning to the previous question, though the slowness of homeostatic adaptation might allow for temporal fluctuations of activity, it does not overcome the conceptual dichotomy between single cell homeostasis and persistent variability in spontaneous activity. Marder and Goaillard speculated that instead of a mechanism controlling activity on a single-cell level, one might consider the possibility of a ``global sensor of network performance to provide error signals for tuning of the network" \cite{Marder_2006}. This concept was used by Sweeney et al. as the basic idea for the introduction of a diffusive, intercellular homeostatic mechanism, which will be introduced in the next section.

@@

\textbf{Synaptic Pruning:} At the same rate of 1/sec, EE synapses below a threshold of 0.000001 mV were removed, thus being added again to the set of ``potential" connections from which the growth process draws new connections.

\textbf{Spike Timing Dependent Plasticity:} An additive STDP rule was used as described e.g. in \cite{Zhang_STDP}. The change of weight between two neurons due to a pre- and postsynaptic spike (i $\rightarrow$ j) is defined as:
\begin{align}
\Delta w_{ji} &= \sum_k \sum_l W(t_j^l - t_i^k) \label{STDP_rule} \\
W(\Delta t) &= A_{+} exp(-\Delta t / \tau_{+}), & \Delta t > 0 \label{STDP_pos} \\
W(\Delta t) &= A_{-} exp(\Delta t / \tau_{-}), & \Delta t < 0 \label{STDP_neg}
\end{align}
Indexes k and l refer to the k-th and l-th pre- and postsynaptic spike respectively. Parameters were chosen to approximate data from \cite{Bi_Poo_STDP} and \cite{Froemke_STDP}, namely $\tau_{+} = 15 ms$, $A_{+} = 15 mV$, $\tau_{-} = 30 ms$ and $A_{-} = -7.5 mV$. However, we used the ``nearest neighbor" approximation for the sake of reduction of computational effort, only calculating the effect of the most recent pre-post pair of spikes for potentiation and post-pre pair for depression, yielding roughly the same value as the full summation due to the fast decay times $\tau_{+}$ and $\tau_{-}$ of the STDP-window.

@@

The new diffusive homeostatic model by Sweeney et al. combined the empirical findings discussed in Section \ref{NO_Experiments_Section} into a set of differential equations:   
\begin{align}
\dot{{Ca^{2+}}_i}(t) &= -\frac{{Ca^{2+}}_i}{\tau_{Ca^{2+}}} + Ca^{2+}_{spike} \sum_{j} \delta(t-t^j_{spike,i}) \label{Ca_dyn}\\
\dot{nNOS_i}(t) &= \frac{1}{\tau_{nNOS}} \left(\frac{{{Ca^{2+}}_i}^3}{{{Ca^{2+}}_i}^3+1} - nNOS_i \right) \label{nNOS_dyn}\\
\dot{NO}(\mathbf{r},t)&=-\lambda NO + D \nabla^2 NO + \sum_{i} \delta^2(\mathbf{r}-\mathbf{r}_{neur,i})\cdot nNOS^i \label{NO_dyn}\\
\dot{V_{t,i}}(t) &= \frac{NO(\mathbf{r}_{neur,i},t)-NO_0}{NO_0\cdot\tau_{V_t}} \label{Theta_dyn}
\end{align}

A depolarization within a nerve cell upon a spike-event $t_{spike}$ causes a fixed inflow of ionic current $Ca^{2+}_{spike}$, which is modeled as an instantaneous increase of the $Ca^{2+}$ concentration. The concentration decays exponentially by a time constant $\tau_{Ca^{2+}}$, see \eqref{Ca_dyn}. Though $Ca^{2+}$ currents can be described in a much more detailed fashion, it can be considered as a reasonable approximation \cite[p.~198-203]{Theor_Neur_Dayan}. The influence of $Ca^{2+}$ onto nNOS was modeled by Sweeney et al. through \eqref{nNOS_dyn} using the Hill equation \cite{Hill_Equ} to model a cooperative binding mechanism. The $nNOS$ production is then fed into the pool of nitric oxide via point sources located at the neurons' positions. An additional decay term was added apart from the inflow and the diffusive term to provide a stable finite $NO$ concentration under constant neuronal activity.

@@

The appropriate choice of $NO_0$ is obviously crucial for the goal of achieving and maintaining a certain level of activity. However, one cannot directly set a parameter of the model to the desired population activity as it was the case for canonical intrinsic homeostasis. One rather needs to determine the average concentration \textit{associated} with the desired activity and set it as a target concentration. Though it is possible to derive this relation in an analytic fashion, we let the system run with the previous homeostatic mechanism, still solving equation \eqref{Ca_dyn}-\eqref{NO_dyn} until a steady mean over the concentrations at the neurons' positions was reached. This mean was then set to be the target concentration and we switched to diffusive homeostasis. Table \ref{Params_IP} summarizes the choice of parameters that were introduced in this section. Diffusion parameters roughly match those measured in experiments \cite{Philippides_2000}.
\begin{table}
\begin{tabular}{|l|l|}
\hline
\textbf{parameter} & \textbf{value} \\
\hline
$\mathrm{r_{IP}}$ & $\mathrm{3\;Hz}$ \\
\hline
$\mathrm{\eta_{IP}}$ & $\mathrm{0.1\;mV}$ \\
\hline
$\mathrm{Ca^{2+}_{spike}}$ & 1 \\ \hline
$\mathrm{\tau_{Ca^{2+}}}$ &  $\mathrm{10\;ms}$ \\
\hline
$\mathrm{\tau_{nNOS}}$ & $\mathrm{100\;ms}$ \\
\hline
$\mathrm{D}$ & default: 10 $\mathrm{\mu m^2 ms^{-1}}$ \\
\hline 
$\mathrm{\lambda}$ & $\mathrm{0.1\;s^{-1}}$ \\
\hline
$\mathrm{\tau_{V_t}}$ & see Section \ref{activ_analys} \\
\hline
\end{tabular}
\caption{Parameters of homeostatic intrinsic plasticity.}
\label{Params_IP}
\end{table}

@@

3.) Dirichlet boundary conditions with $NO = NO_{bound}$ at the boundaries.\\
Neumann boundary conditions were used for most of the simulations if not explicitly marked differently. This decision relates to the previously described mechanism of synaptic growth: Neurons placed close to the edge of the sheet have a lower connection probability due to the absence of neighboring neurons in the direction perpendicular to the close-by border. It therefore models the synaptic growth within a square ``cutout" of neural tissue. The Neumann boundary condition fits into this picture since it allows a zero-flux condition at the borders. This is a reasonable assumption, because NO molecules cannot diffuse out of the tissue (unless they were placed in a fluid surrounding).

Equation \eqref{NO_dyn} describes the influx of NO as a sum of scaled and spatially shifted Dirac functions. Apart from the question whether this source term results in a well defined, finite solution at the neurons' positions (see Section \ref{Matrix_Diff_Model_Section}), it can only be modeled to a certain degree of accuracy depending on the resolution of the numeric grid. In practice, we approximated the point sources of NO as an insertions at individual grid cells at a rate of $nNOS_i(t) / h^2$, where the normalizing divisor $h^2$ ensured the desired total influx per neuron. This numeric implementation required two additional constraints: First, all random neuron positions were confined to integer multiples of $h$ in x- and y-direction. Second, to avoid redundancy and for physiological reasons, each grid cell could only hold one neuron at maximum. Both conditions combined led to an iterative generation of positions where for each neuron a random number generator produced pairs of integers $(n_x,n_y)$, each within $[0,N]$, until an unoccupied pair of integers was found and occupied, moving on to the next neuron. Figure \ref{diff_test_plot} shows an example of the resulting NO density throughout a simulation. 
\begin{figure}[h!]
%% density_plot.py
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/density_plots/density_plot.png}
\end{center}
\caption{Example of NO-diffusion with 400 point sources of excitatory neurons.}
\label{diff_test_plot}
\end{figure}

\newpage
\section{Results} \label{results}
We first present the results of a comparison between features of the network activity in both variants in section \ref{activ_analys}, since our main goal of the implementation of diffusive homeostasis was to allow the network to develop a broader distribution of firing rates across excitatory neurons compared to the original version of homeostasis. This is followed by an analytic discussion of an instability we observed within the diffusive homeostatic feedback loop, see Section \ref{theor_osc}. Furthermore, we compare topological features of the network under the influence of diffusive and non-diffusive homeostasis in section \ref{Network_Topology_Section}.

@@

where $\tau_e$ and $\tau_i$ are time constants representing the rapidity of firing rate adaptation, $\phi_e$ and $\phi_i$ the respective transition functions between synaptic input and firing rate and $W_{xy}$ the mean synaptic weight from population x to population y. A persistent oscillation then corresponds to a stable limit cycle in the 2-dimensional phase plane. Such a limit cycle exists inside a region $R$ if $R$ contains no fixed points and if any trajectory whose starting points lies within $R$ is confined to $R$, according to the Poincaré-Bendixson theorem. The second condition is equivalent to the condition that for all points $\mathbf{r}_{edge}$ on the edge of $R$, the vector field $\dot{\mathbf{r}}_{edge}$ is facing into $R$, see \cite[p.~248]{Dyn_Sys_Hirsch}. This case occurs if the system has a locally unstable fixed point, but a non-linearity that prevents an infinite deviation from the fixed point.

@@

{Ca^{2+}}^3(t) &= \left[ Ca^{2+}_{spike} \sum_{i} \theta(t-t^i_{spike}) exp(-(t-t^i_{spike})/\tau_{Ca^{2+}}) \right]^3 \\
&\approx {Ca^{2+}_{spike}}^3 \sum_{i} \theta(t-t^i_{spike}) exp(-3(t-t^i_{spike})/\tau_{Ca^{2+}})

@@

Returning to the question of how to describe NO synthesis in a simplified form, we note that according to equation \eqref{nNOS_single_sol_approx2}, a single spike causes the release of NO by an amount of $\frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$, which makes the mean rate of NO production over time simply $<nNOS> = <r> \cdot \frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$. However, it is not sufficient to simply propose $nNOS_i(t) = r_i(t) \cdot \frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$. A single cell fires in the range of $\mathrm{3\,Hz}$, which is - at least on the time scale of the observed oscillations - not enough to assign an instantaneous firing rate.
\newpage
We introduced an idealized picture of diffusive homeostasis to deal with this, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses $nNOS_{total}(t) \equiv \sum_{i} nNOS^i(t)$ and the decay term $-\lambda NO$. , $nNOS_{total}$ must be divided by the area $L^2$ of the tissue to account for the spread of NO. Furthermore, we introduce a random fluctuation term $\sigma_{NO} \xi(t)$ that accounts for local, momentary deviations from the mean. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{nNOS_{total}(t)}{L^2} + \sigma_{NO} \xi(t)\;.
\label{NO_dyn_approx}
\end{equation}
The amount of noise in the system is a hard-to-predict quantity since it depends on the spatial structure on the diffusive lattice, giving neurons that are further separated less impact onto each other. We tried to estimate the noise level based on the same instant-spread-simplification made in Equation \eqref{NO_dyn_approx} to describe the overall NO-dynamics. This led to an underestimation of the total energy within the power spectrum in the final result. However, noise level only acts as a global scaling factor of the shape of the power spectrum as long as we assume the noise in the system to be approximately white and was thus left open as a model fitting parameter.

@@

Figure \ref{Pow_Spec_Theta_vs_Analytic} shows the result of \eqref{Pow_Spec_Theta_Final}. Choosing noise amplitude to be the only free parameter of the model gave a rather unsatisfying result. Our approximation underestimated the amount of damping within the system (which controls the ``peakiness" of the spectrum), as well as predicting a slightly smaller preferred amplitude. The preferred amplitude is controlled by $\frac{\gamma \alpha N_{exc.}}{L^2}$ and $\frac{1}{NO_0\tau_{V_t}}$ while the broadness of the spectrum is mostly controlled by the damping term $-\lambda$. Since the latter term was taken unaltered from the original system (apart from a shift in phase space), we added $-\lambda$ and $\alpha$ as free fitting parameters and managed to get a very good fit to the simulation data, see Figure \ref{Pow_Spec_Theta_vs_Analytic}.   
\begin{figure}
%% plot_pow_spec_analytic.py
\begin{center}
\includegraphics[width=\textwidth]{../../plots/power_spec/power_spectrum_analytic.png}
\end{center}
\caption{Power Spectrum of $\mathrm{\langle V_t \rangle}$ (blue) and analytic predictions based on Equation \eqref{Pow_Spec_Theta_Final}. Red curve was fitted by tweaking only the overall amplitude $\mathrm{\sigma_{NO}}$ in \eqref{Pow_Spec_Theta_Final} ($\mathrm{\alpha = -0.789 \,Hz/mV}$, as in Figure \ref{thresh_r_linfit}). Green curve is a fit achieved by adjusting $\mathrm{\sigma_{NO}}$ and $\mathrm{\lambda}$ in \eqref{Pow_Spec_Theta_Final} ($\mathrm{\lambda = 0.845 \,s^{-1}}$). Black curve was fitted by means of $\mathrm{\sigma_{NO}}$, $\mathrm{\lambda}$ and $\mathrm{\alpha}$ ($\mathrm{\lambda = 0.592 \,s^{-1}}$, $\mathrm{\alpha=-0.989 \,Hz/mV}$).}
\label{Pow_Spec_Theta_vs_Analytic}
\end{figure}
The fact that the value for $\mathrm{\alpha}$ that was determined by means of the linear fit in Figure \ref{thresh_r_linfit} leads to an underestimation of the preferred frequency can be partially explained by the linear regression, which overestimates the importance of a number of high-threshold-outliers. This leads to a shallower slope than what would be appropriate for the working point of oscillatory behavior. Another possible explanation could be that ``instant spread" is only an approximation. Rather, one could picture the nitrous oxide just to fill a fraction of the available space on a short time scale by replacing $L^2$ by a smaller area. This leads to a stronger local increase of concentration, which eventually leads to faster oscillations. In either case, both parameters have the same effect onto the shape of the power spectrum since these parameters appear as a fraction $\alpha/L^2$ in \eqref{Pow_Spec_Theta_Final}.

@@

From a more general perspective one could argue that a mechanism binding firing rates to spatial constraints suppresses the ability of the system to adapt to a changing environment since it essentially ``freezes" the state of activity. Experiments have shown that neurons' firing rates are indeed correlated under a changing environment/task: Neurons with a certain baseline-firing rate are very likely to have a similar firing rate while performing a task \cite{Buzsaki_Fir_Rates_2014}. While homeostatic fixation might not be problematic in the case of transient changes of stimuli, a spatial fixation of firing rates would presumably contradict the fact that long-term learning processes and plasticity lead to persistent changes in activity patterns \cite{Lever_Long_Term_Plast_2002}.

@@

Taking a closer look at the distributions depicted in Figure \ref{Weight_Dist} we noted that the increased width of the distribution is largely due to to an increment on the left half of the curve. Since new weights are initialized at $\mathrm{10^{-4}\, mV}$ this might indicate that some weights are drifting much more slowly from their value of initialization compared to other weights. These ``slow drifters" are presumably connecting neurons both spiking at a below-average firing rate.
This provides an explanation why this effect is more present in the case of diffusive homeostasis since it causes a portion of neurons to have rather low firing rates. On the other hand, for non-diffusive homeostasis, one can assume that the amount of STDP-fluctuations is approximately equal for all recurrent excitatory weights because of the fixation of firing rates.
